Efficient Computation of Extended Sources

### 2021-03-28 repro/papers/sseprecond/paper.tex
snapshot

\section{Introduction}
Full Waveform Inversion (FWI) can be described in terms of 
\begin{enumerate}
\item a linear wave operator $L[c]$, depending on an array of space-dependent coefficients $c$ and acting on wavefields $u$ vanishing in negative time:
\begin{equation}
\label{eqn:init}
u \equiv 0, t \ll 0; 
\end{equation}
\item a trace sampling operator $P$ acting on wavefields and producing data traces;
\item and a source function (of space and time) $f$ representing energy input to the system. 
\end{enumerate}
The basic FWI problem is: given data $d$, find $c$ so that 
\begin{equation}
\label{eqn:fwi}
Pu \approx d \mbox{ and } L[c]u = f.
\end{equation}

The energy source $f$ may also be largely undetermined, apart from some known characteristics such as localization in space and/or time. In fact, additional source degrees of freedom, beyond those needed to describe physically realized sources, may be useful in rendering the FWI problem \ref{eqn:fwi} more amenable to numerical solution, via so-called extended modeling (see \cite{geoprosp:2008}, \cite{LeeuwenHerrmannWRI:13}, \cite{HuangNammourSymesDollizal:SEG19}, and many references cited there). Therefore it is natural to view $f$ as also an unknown in formulating the problem \ref{eqn:fwi} via nonlinear least squares:
%A typical nonlinear least squares formulation is:
%\begin{equation}
%\label{eqn:ols}
%\mbox{choose } c \mbox{ to minimize } \|PL[c]^{-1}f -d \|^2.
%\end{equation}

%As is well-known, local optimization methods are the only feasible approach given the dimensions of a typical instance of \ref{eqn:fwi}, and those have a tendency to stall due to ``cycle-skipping''. Source extension is one approach to avoiding this problem. It consists in imposing the wave equation as a soft as opposed to hard constraint, by allowing the source to have more degrees of freedom than is permitted by a faithful  model model of the seismic experiment, and constraining these additional degrees of freedom by means of an additive penalty modifying the probem \ref{eqn:ols}:
\begin{equation}
\label{eqn:esi}
\mbox{choose } c, f \mbox{ to minimize } \|PL[c]^{-1}f -d \|^2 + \alpha^2 \|Af\|^2 
\end{equation}
The operator $A$ penalizes deviation from known (or assumed) characteristics of the source function - its null space consists of feasible source models.

This paper presents a numerically efficient approach to solving the {\em source subproblem}
\begin{equation}
\label{eqn:esis}
\mbox{given } c, \mbox{ choose } f \mbox{ to minimize } \|PL[c]^{-1}f -d \|^2 + \alpha \|Af\|^2 
\end{equation}

@@

The source subproblem is a necessary step in solving the overall FWI problem \ref{eqn:fwi}. In particular, the nonlinear least squares realization \ref{eqn:esi} of source-extended FWI is naturally approached via the {\em variable projection method} \cite[]{GolubPereyra:73,GolubPereyra:03,vanLeeuwenMulder:09,Rickett:SEG12}, in which the minimum over $f$ in \ref{eqn:esis} is treated as a function of $c$, which is in turn minimized. Variable projection thus treats the source subproblem as an inner problem, which must be solved for an estimate of $f$ at every step of an iterative method for the outer problem of minimization over $c$. Efficiency in solving the source subproblem \ref{eqn:esis} is critical to efficient implementation of variable projection for solving the regularized FWI problem \ref{eqn:esi}.

This paper restricts the choice of wave physics embodied in $L[c]$ to linear acoustics.  The principal characteristics assumed of sources are localization and isotropy. That is, physical sources will be assumed to be isotropic point radiators, described by the functional form
\begin{equation}
\label{eqn:ptsrc}
f_{\rm pt}(\bx,t;\bx_s) = w(t)\delta(\bx-\bx_s). 
\end{equation}
Source locations $\bx_s$ are assumed known (as is natural for active source methods), and the source {\em wavelet} $w(t)$ is to be found as part of the solution of \ref{eqn:esi}. A convenient choice of penalty operator $A$ is then multiplication by the {\em offset} $|\bx-\bx_s|$: its null space consists precisely of distributions of the form \ref{eqn:ptsrc}.

@@

\section{Operators and Adjoints}

For acoustic wave physics, the coefficient vector $c$ is the pair $c=(\kappa,\beta)$ of bulk modulus $\kappa$ and density $\rho$, and the state vector $u=(p,\bv)^T$ consists of pressure $p$ (a scalar space-time field) and particle velocity $\bv$ (a vector space-time field). The wave operator $L[c]$ is given by the linear acoustics system:
\begin{equation}
\label{eqn:aweop}
L[c]u = 
\left(
\begin{array}{c}
\frac{1}{\kappa}\frac{\partial p}{\partial t}  + \nabla \cdot \bv, \\
\rho\frac{\partial \bv}{\partial t} + \nabla p.
\end{array}
\right) 
\end{equation}
That is,
\begin{equation}
  \label{eqn:awemat}
  L[c] = \left(
    \begin{array}{cc}
      \frac{1}{\kappa}\frac{\partial}{\partial t} & \nabla \cdot \\
      \nabla & \rho \frac{\partial}{\partial t}
    \end{array}
  \right)
\end{equation}

The space dimension is $d$. Most of what follows is valid for any $d >0$. Examples later in this paper will use $d=2$ for computationally convenience.

As mentioned earlier, in this paper the coefficients $c$ is regarded as defined throughout space $\bR^d$, the state vector $u$ throughout space-time $\bR^{d+1}$.

The idealized ``crosswell'' configuration involves sources confined to $z=z_s$. I will mostly represent sources as constitutive law defects, that is, right-hand sides in the system $L[c]u=f$ of the form $f(\bx,t) = (h_s(x,y,t)\delta(z-z_s), {\bf 0})^T$. With the choice $L[c]$ given in \ref{eqn:aweop}, the causal wave system $L[c]u^+=f$ takes the form
\begin{eqnarray}
\label{eqn:awep}
\frac{1}{\kappa}\frac{\partial p^{+}}{\partial t} & = & - \nabla \cdot \bv^{+} +
h_s \delta(z-z_s), \nonumber \\
\rho\frac{\partial \bv^{+}}{\partial t} & = & - \nabla p^{+},\nonumber \\
p^{+} & =& 0 \mbox{ for } t \ll 0,\nonumber\\ 
\bv^{+} & = & 0 \mbox{ for } t \ll 0.
\end{eqnarray}

Similarly, the anti-causal wave system $L[c]u^-=f$ takes the form
\begin{eqnarray}
 \label{eqn:awem}
\frac{1}{\kappa}\frac{\partial p^{-}}{\partial t} & = & - \nabla \cdot \bv^{-} +
h_s \delta(z-z_s), \nonumber \\
\rho \frac{\partial \bv^{-}}{\partial t} & = & - \nabla p^{-},\nonumber \\
p^{-} & =& 0 \mbox{ for } t \gg 0,\nonumber\\ 
\bv^{-} & = & 0 \mbox{ for } t \gg 0.
\end{eqnarray}
Sampling the causal pressure field $p^{+}$ on the surface $z=z_r$ defines the causal forward modeling operator $S^{+}_{z_s,z_r}[c]$:
\begin{equation}
  \label{eqn:fwd}
  S^{+}_{z_s,z_r}[c]h_s = p^+|_{z=z_r}
\end{equation}
The anti-causal forward modeling operator $S^-_{z_s,z_r}$ has a similar definition.
The subscript signifies that sources are located on $z=z_s$, the receivers on $z=z_r$. It's necessary to include this information in the notation, as versions of $S$ with sources and receivers in several locations will be needed in the discussion below.

The adjoint of $S^+_{z_s,z_r}[c]$ can be computed
by a variant of the adjoint state method, in this case a by-product of the conservation of energy. Suppose that $p^-,\bv^-$ solve \ref{eqn:awem} with $h_s\delta(z-z_s)$ replaced by $ h_r \delta(z-z_r)$. Then
\[
0 = 
\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  
\rho \bv^+ \cdot \bv^- \right)|_{t \rightarrow \infty}
-
\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  \rho \bv^+ \cdot \bv^- \right)|_{t \rightarrow -\infty}
\]
\[
= 
\int_{-\infty}^{\infty} \,dt\, \frac{d}{dt}\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  \rho \bv^+ \cdot \bv^- \right)
\]
\[
= 
\int_{-\infty}^{\infty} \,dt\, \left(\int\, dx\,dy\,dz\, \frac{1}{\kappa} \frac{\partial p^+}{\partial t} p^- +  p^+ \frac{1}{\kappa}\frac{\partial p^-}{\partial t} \right.
\]
\[
+
\left. \rho \frac{\partial \bv^+}{\partial t} \cdot \bv^- + \rho \bv^+ \cdot \frac{\partial \bv^-}{\partial t} \right)
\]
\[
= 
\int_{-\infty}^{\infty} \,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv^+ + 
 h_s \delta(z-z_s)\right) p^- + p^+ \left(- \nabla \cdot \bv^- + 
 h_r \delta(z-z_r)\right) \right.
\]
\[
+
\left.  (- \nabla p^+) \cdot \bv^- + \bv^+ \cdot (-\nabla p^-) \right)
\]
\[
= 
\int_{-\infty}^{\infty}\,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv^+ + 
 h_s \delta(z-z_s)\right) p^- + p^+ \left(- \nabla \cdot \bv^- + 
 h_r \delta(z-z_r)\right) \right.
\]
\[
+
\left.  p^+ (\nabla \cdot \bv^-) + (\nabla \cdot \bv^+) p^- \right)
\]
after integration by parts in the last two terms. Most of what is left cancels, leaving 
\[
0 = \int_{-\infty}^{\infty}\,dt\,dx\,dy\, h_sp^-|_{z=z_s} +  h_rp^+|_{z=z_r} = \langle h_s, p^-|_{z=z_s}\rangle + \langle h_r, S[c]^+_{z_s,z_r}[c]h_s \rangle
\]
whence
\begin{equation}
\label{eqn:sadj}
 S^+_{z_s,z_r}[c]^T h_r = -p^-|_{z=z_s}
\end{equation}

The systems \ref{eqn:awep} and \ref{eqn:awem} differ only in the direction of time evolution (causal vs. anti-causal). Define $R$ to be the {\em time-reversal operator} on functions of space-time,
$Rf(\bx,t) = f(\bx,-t)$, and ${\cal R}$ to be the {\em acoustic field time-reversal operator} 
\begin{equation}
  \label{eqn:trdef}
  {\cal R} \left(
    \begin{array}{c}
      p\\
      \bv
    \end{array}
  \right) =
  \left(
    \begin{array}{c}
      Rp\\
      -R\bv
    \end{array}
  \right)
\end{equation}
Then ${\cal R}(p^-,\bv^-)$ solves \ref{eqn:awep} with $h_s\delta(z-z_s)$ replaced by $-Rh_r\delta(z-z_r)$. That is,
\begin{equation}
  \label{eqn:trsadj}
  Rp^-|_{z=z_s} = -S^+_{z_r,z_s}[c]Rh_r
\end{equation}
Since $R^2 = I$ and ${\cal R}^2 = I$, the identities \ref{eqn:sadj} and \ref{eqn:trsadj} imply that
\begin{equation} 
  \label{eqn:trtr}
  S^+_{z_s,z_r}[c]^T = RS^+_{z_r,z_s}[c]R.
\end{equation}

@@

A particular case of the relation \ref{eqn:trtr} is important in its own right, when $z_r=z_s$. Define surface source-to-pressure operator $\Lambda_{z_s}$ by
\begin{equation}
  \label{eqn:ntoddef}
  \Lambda_{z_s} = S^+_{z_s,z_s}.
\end{equation}
Then \ref{eqn:trtr} reads
\begin{equation}
  \label{eqn:ntodtr}
  \Lambda_{z_s}^T = R\Lambda_{z_s}R.
\end{equation}
This operator is closely related to the operator mapping $\lim_{z\rightarrow z_s^+} \bv_z^+$ to $\lim_{z\rightarrow z_s^+}  p^+$, often called the {\em Neumann-to-Dirichlet} operator. The connection is explained below.

The quadratic form defined by $\Lambda_{z_s}$ has fundamental physical significance. Suppose that the surface source $h_s$ vanishes outside of a bounded time interval $[t_{\rm min},t_{\rm max}]$. Then the causal acoustic field $(p^+,\bv^+)$, solution of the system \ref{eqn:awep}, vanishes for $t<t_{\rm min}$. Define the total acoustic energy $E(t)$ at time $t$ by
\begin{equation}
  \label{eqn:defae}
  E(t) = \frac{1}{2} \int \,d\bx \, \left(\frac{(p^+)^2}{\kappa} + \rho |\bv^+|^2\right)
\end{equation}
Then since $E(t_{\rm min})=0$,
\[
  E(t_{\rm max}) = \int_{t_{\rm min}}^{t_{\rm max}} \frac{dE}{dt}
\]
\[
  = \int_{t_{\rm min}}^{t_{\rm max}} \int\,d\bx\, \left(p^+ \frac{1}{\kappa}\frac{\partial p^+}{\partial t} + \bv^+ \cdot \rho \frac{\partial \bv^+}{\partial t}\right)
\]
\[
  =\int_{t_{\rm min}}^{t_{\rm max}} \int\,d\bx\,\left(p^+(-\nabla \cdot \bv^{+} + h_s\delta(z-z_s)) + \bv^+ \cdot \rho \frac{\partial \bv^+}{\partial t}\right)
\]
\[
  \int_{t_{\rm min}}^{t_{\rm max}}\int\,d\bx\, \left(\left(\nabla p^+\rho \frac{\partial \bv^+}{\partial t}\right) + p^+h_s\delta(z-z_s)\right)
\]
\[
  =  \int_{t_{\rm min}}^{t_{\rm max}}\int\,dxdy h_s p^+|_{z=z_s} = \langle h_s, \Lambda_{z_s} h_s \rangle.
\]
Since the source $h_s\delta(z-z_s)$ no longer acts on the fluid for $t >t_{\rm max}$, $E_{\rm tot}= E(t_{\rm max})$ is the total energy transferred from the source to the fluid, and
\begin{equation}
  \label{eqn:surfenergy}
  E_{\rm tot} = \langle h_s, \Lambda_{z_s} h_s \rangle.
\end{equation}
We can conclude from identity \ref{eqn:surfenergy} that the symmetric part of $\Lambda_{z_s}$ is positive semidefinite:
\begin{equation}
  \label{eqn:lamsd}
  \frac{1}{2}(\Lambda_{z_s}^T + \Lambda_{z_s}) \ge 0.
\end{equation}

Observe that the causal acoustic system \ref{eqn:awep} implies directly that
\begin{equation}
  \label{eqn:sourcevjump}
  h_s= [v_z]|_{z_s}.
\end{equation}

>>>>

\section{Introduction}
Full Waveform Inversion (FWI) can be described in terms of 
\begin{enumerate}
\item a linear wave operator $L[\bf{c}]$, depending on an array of space-dependent coefficients $\bf{c}$ and acting on wavefields $\bu$ vanishing in negative time:
\begin{equation}
\label{eqn:init}
\bu \equiv 0, t \ll 0; 
\end{equation}
\item a trace sampling operator $P$ acting on wavefields and producing data traces;
\item and a source function (of space and time) $\bff$ representing energy input to the system. 
\end{enumerate}
The basic FWI problem is: given data $d$, find $\bf{c}$ so that 
\begin{equation}
\label{eqn:fwi}
P\bu \approx d \mbox{ and } L[\bf{c}]\bu = \bff.
\end{equation}

The energy source $\bff$ may also be largely undetermined, apart from some known characteristics such as localization in space and/or time. In fact, additional source degrees of freedom, beyond those needed to describe physically realized sources, may be useful in rendering the FWI problem \ref{eqn:fwi} more amenable to numerical solution, via so-called extended modeling (see \cite{geoprosp:2008}, \cite{LeeuwenHerrmannWRI:13}, \cite{HuangNammourSymesDollizal:SEG19}, and many references cited there). Therefore it is natural to view $\bff$ as also an unknown in formulating the problem \ref{eqn:fwi} via nonlinear least squares:
%A typical nonlinear least squares formulation is:
%\begin{equation}
%\label{eqn:ols}
%\mbox{choose } c \mbox{ to minimize } \|PL[c]^{-1}f -d \|^2.
%\end{equation}

%As is well-known, local optimization methods are the only feasible approach given the dimensions of a typical instance of \ref{eqn:fwi}, and those have a tendency to stall due to ``cycle-skipping''. Source extension is one approach to avoiding this problem. It consists in imposing the wave equation as a soft as opposed to hard constraint, by allowing the source to have more degrees of freedom than is permitted by a faithful  model model of the seismic experiment, and constraining these additional degrees of freedom by means of an additive penalty modifying the probem \ref{eqn:ols}:
\begin{equation}
\label{eqn:esi}
\mbox{choose } \bf{c}, \bff \mbox{ to minimize } \|PL[\bf{c}]^{-1}\bff -d \|^2 + \alpha^2 \|A\bff\|^2 
\end{equation}
The operator $A$ penalizes deviation from known (or assumed) characteristics of the source function - its null space consists of feasible source models.

This paper presents a numerically efficient approach to solving the {\em source subproblem}
\begin{equation}
\label{eqn:esis}
\mbox{given } \bf{c}, \mbox{ choose } \bff \mbox{ to minimize } \|PL[\bf{c}]^{-1}\bff -d \|^2 + \alpha \|A\bff\|^2 
\end{equation}

@@

The source subproblem is a necessary step in solving the overall FWI problem \ref{eqn:fwi}. In particular, the nonlinear least squares realization \ref{eqn:esi} of source-extended FWI is naturally approached via the {\em variable projection method} \cite[]{GolubPereyra:73,GolubPereyra:03,vanLeeuwenMulder:09,Rickett:SEG12}, in which the minimum over $\bff$ in \ref{eqn:esis} is treated as a function of $\bf{c}$, which is in turn minimized. Variable projection thus treats the source subproblem as an inner problem, which must be solved for an estimate of $\bff$ at every step of an iterative method for the outer problem of minimization over $\bf{c}$. Efficiency in solving the source subproblem \ref{eqn:esis} is critical to efficient implementation of variable projection for solving the regularized FWI problem \ref{eqn:esi}.

This paper restricts the choice of wave physics embodied in
$L[\bf{c}]$ to linear acoustics.  Thus $\bf{c}=(\kappa,\rho)$ is the
pair (bulk modulus, density) of positive functions of spatial
position. The principal characteristics assumed of sources $\bff=$ are localization and isotropy. That is, physical sources will be assumed to be isotropic point radiators, described by the functional form
\begin{equation}
\label{eqn:ptsrc}
f_{\rm pt}(\bx,t;\bx_s) = w(t)\delta(\bx-\bx_s). 
\end{equation}
Source locations $\bx_s$ are assumed known (as is natural for active source methods), and the source {\em wavelet} $w(t)$ is to be found as part of the solution of \ref{eqn:esi}. A convenient choice of penalty operator $A$ is then multiplication by the {\em offset} $|\bx-\bx_s|$: its null space consists precisely of distributions of the form \ref{eqn:ptsrc}.

@@

\section{Surface Sources and Solutions}

For acoustic wave physics, the coefficient vector $c$ is the pair $\bf{c}=(\kappa,\rho)$ of bulk modulus $\kappa$ and density $\rho$, and the state vector $\bu=(p,\bv)^T$ consists of pressure $p$ (a scalar space-time field) and particle velocity $\bv$ (a vector space-time field). The wave operator $L[\bf{c}]$ is given by the linear acoustics system:
\begin{equation}
\label{eqn:aweop}
L[c]u = 
\left(
\begin{array}{c}
\frac{1}{\kappa}\frac{\partial p}{\partial t}  + \nabla \cdot \bv, \\
\rho\frac{\partial \bv}{\partial t} + \nabla p.
\end{array}
\right) 
\end{equation}
That is,
\begin{equation}
  \label{eqn:awemat}
  L[c] = \left(
    \begin{array}{cc}
      \frac{1}{\kappa}\frac{\partial}{\partial t} & \nabla \cdot \\
      \nabla & \rho \frac{\partial}{\partial t}
    \end{array}
  \right)
\end{equation}

Most of what follows is valid for any space dimension $n >0$. Examples
later in this paper will use $n=2$ for computationally convenience. In
the mathematical development that follows, I will set $n=3$.

As mentioned earlier, in this paper the coefficients $\bf{c}$ is regarded as defined throughout space $\bR^d$, the state vector $\bu$ throughout space-time $\bR^{d+1}$.

The idealized ``crosswell'' configuration involves sources confined to
$z=z_s$. Such sources are combinations of constitutive law defects and
loads normal to the surface, that is, right-hand sides in the system
$L[c]\bf{u}=\bf{f}$ of the form $\bf{f}(\bx,t) =
(h_s(x,y,t)\delta(z-z_s), f_s(x,y,t)\bf{e}_z\delta(z-z_s))^T$ for scalar
defect $h_s$ and normal force $f_s$ and $\bf{e}_z=(0,0,1)^T$. With the choice $L[\bf{c}]$ given in \ref{eqn:aweop}, the causal wave system $L[\bf{c}]\bu^+=\bff$ takes the form
\begin{eqnarray}
\label{eqn:awep}
\frac{1}{\kappa}\frac{\partial p^{+}}{\partial t} & = & - \nabla \cdot \bv^{+} +
h_s \delta(z-z_s), \nonumber \\
\rho\frac{\partial \bv^{+}}{\partial t} & = & - \nabla p^{+} + f_s\bf{e}_z \delta(z-z_s),\nonumber \\
p^{+} & =& 0 \mbox{ for } t \ll 0,\nonumber\\ 
\bv^{+} & = & 0 \mbox{ for } t \ll 0.
\end{eqnarray}

Similarly, the anti-causal wave system $L[c]u^-=f$ takes the form
\begin{eqnarray}
 \label{eqn:awem}
\frac{1}{\kappa}\frac{\partial p^{-}}{\partial t} & = & - \nabla \cdot \bv^{-} +
h_s \delta(z-z_s), \nonumber \\
\rho \frac{\partial \bv^{-}}{\partial t} & = & - \nabla p^{-} + f_s\bf{e}_z \delta(z-z_s),\nonumber \\
p^{-} & =& 0 \mbox{ for } t \gg 0,\nonumber\\ 
\bv^{-} & = & 0 \mbox{ for } t \gg 0.
\end{eqnarray}

\section{Operators and Adjoints}
All of the operators discussed in this section depend on the
coefficient array $\bf{c}$, which I will suppress from the notation.
Denote by $P_s,P_r$ the trace (sampling) operators on $z=z_s$, $z=z_r$
respectively. The causal and anticausal modeling operators
${\cal S}^{+}_{z_s,z_r}$ and ${\cal S}^-_{z_s,z_r}$ are defined in
terms of the solutions $(p^{\pm},\bv^{\pm})$ of the systems
\ref{eqn:awep} and \ref{eqn:awem} by
\begin{eqnarray}
  {\cal S}^{+}_{z_s,z_r}(h_s,f_s) & = & (P_rp^+,P_r v_z^+)|_{z=z_r},
                                        \nonumber\\
  {\cal S}^{-}_{z_s,z_r}(h_s,f_s) & = & (P_rp^-,P_r v_z^-)|_{z=z_r}.
  \label{eqn:fwd}
\end{eqnarray}
The subscript signifies that sources are located on $z=z_s$, the
receivers on $z=z_r$. It is necessary to include this information in
the notation, as versions of ${\cal S}^{\pm}$ with sources and receivers in
several locations will be needed in the discussion below.

The adjoint of ${\cal S}^+_{z_s,z_r}$ can be computed by a variant of
the adjoint state method, in this case a by-product of the
conservation of energy. Suppose that $p^-,\bv^-$ solve \ref{eqn:awem}
with $(h_s,f_s\bf{e}_z)\delta(z-z_s)$ replaced by
$ (h_r,f_r\bf{e}_z) \delta(z-z_r)$. Then
\[
0 = 
\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  
\rho \bv^+ \cdot \bv^- \right)|_{t \rightarrow \infty}
-
\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  \rho \bv^+ \cdot \bv^- \right)|_{t \rightarrow -\infty}
\]
\[
= 
\int_{-\infty}^{\infty} \,dt\, \frac{d}{dt}\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  \rho \bv^+ \cdot \bv^- \right)
\]
\[
= 
\int_{-\infty}^{\infty} \,dt\, \left(\int\, dx\,dy\,dz\, \frac{1}{\kappa} \frac{\partial p^+}{\partial t} p^- +  p^+ \frac{1}{\kappa}\frac{\partial p^-}{\partial t} \right.
\]
\[
+
\left. \rho \frac{\partial \bv^+}{\partial t} \cdot \bv^- + \rho \bv^+ \cdot \frac{\partial \bv^-}{\partial t} \right)
\]
\[
= 
\int_{-\infty}^{\infty} \,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv^+ + 
 h_s \delta(z-z_s)\right) p^- + p^+ \left(- \nabla \cdot \bv^- + 
 h_r \delta(z-z_r)\right) \right.
\]
\[
+
\left.  (- \nabla p^++f_s\bf{e}_z) \cdot \bv^- + \bv^+ \cdot (-\nabla
  p^- + f_r \bf{e_z}) \right)
\]
\[
= 
\int_{-\infty}^{\infty}\,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv^+ + 
 h_s \delta(z-z_s)\right) p^- + p^+ \left(- \nabla \cdot \bv^- + 
 h_r \delta(z-z_r)\right) \right.
\]
\[
+
\left.  p^+ (\nabla \cdot \bv^-) + (\nabla \cdot \bv^+) p^- 
  +f_s \delta(z-z_s) v_z^- + v_z^+f_r \delta(z-z_r) \right)
\]
after integration by parts in the last two terms. Most of what is left cancels, leaving 
\[
0 = \int_{-\infty}^{\infty}\,dt\,dx\,dy\, (h_sP_sp^-+f_zP_sv_z^-) +
( h_rP_rp^++f_rP_rv_z^+) = \langle (h_s,f_s), {\cal S}^-(h_r,f_r) \rangle+ \langle (h_r,f_r), {\cal S}^+_{z_s,z_r}(h_s,f_s) \rangle
\]
whence
\begin{equation}
\label{eqn:sadj}
 ({\cal S}^+_{z_s,z_r})^T = -{\cal S}^{-}_{z_r,z_s}.
\end{equation}

The systems \ref{eqn:awep} and \ref{eqn:awem} differ only in the direction of time evolution (causal vs. anti-causal). Define $R$ to be the {\em time-reversal operator} on functions of space-time,
$Rf(\bx,t) = f(\bx,-t)$, and ${\cal R}$ to be the {\em acoustic field time-reversal operator} 
\begin{equation}
  \label{eqn:trdef}
  {\cal R} \left(
    \begin{array}{c}
      p\\
      \bv
    \end{array}
  \right) =
  \left(
    \begin{array}{c}
      Rp\\
      -R\bv
    \end{array}
  \right)
\end{equation}
Then ${\cal R}(p^-,\bv^-)$ solves \ref{eqn:awep} with
$(h_s\delta(z-z_s),f_s\bf{e_z}\delta(z-z_s))$ replaced by
$(-Rh_r\delta(z-z_r), Rf_r\bf{e_z}\delta(z-z_r))$. That is,
\begin{equation}
  \label{eqn:trsadj}
  {\cal R}{\cal S}^- = -{\cal S}^+_{z_r,z_s}{\cal R}
\end{equation}
Since $R^2 = I$ and ${\cal R}^2 = I$, the identities \ref{eqn:sadj} and \ref{eqn:trsadj} imply that
\begin{equation} 
  \label{eqn:trtr}
 ({\cal S}^+_{z_s,z_r})^T = {\cal R}{\cal S}_{z_r,z_s}^+{\cal R}=
 -{\cal S}^-_{z_r,z_s}.
\end{equation}

@@

A particular case of the relation \ref{eqn:trtr} is important in its own right, when $z_r=z_s$. Define surface source-to-pressure operator $\Lambda_{z_s}$ by
\begin{equation}
  \label{eqn:ntoddef}
  \Lambda^{\pm}_{z_s} = S^{\pm}_{z_s,z_s}.
\end{equation}
Then \ref{eqn:trtr} reads
\begin{equation}
  \label{eqn:ntodtr}
  (\Lambda^+_{z_s})^T = R\Lambda_{z_s}^+R = -\Lambda^-_{z_s}
\end{equation}
This operator is closely related to the operator mapping $\lim_{z\rightarrow z_s^{\pm}} \bv_z^+$ to $\lim_{z\rightarrow z_s^{\pm}}  p^+$, often called the {\em Neumann-to-Dirichlet} operator. The connection is explained below.

The quadratic form defined by $\Lambda^{\pm}_{z_s}$ has fundamental physical
significance. Define the total acoustic energy $E^{\pm}(t)$ of the field $(p^{\pm},\bv^{\pm})$, at time $t$ by
\begin{equation}
  \label{eqn:defae}
  E^{\pm}(t) = \frac{1}{2} \int \,d\bx \, \left(\frac{(p^{\pm})^2}{\kappa} + \rho |\bv^{\pm}|^2\right)
\end{equation}
Then for $t_{\min}<t_{\max}$,
\[
  E^{\pm}(t_{\rm max})-E^{\pm}(t_{\rm min}) = \int_{t_{\rm min}}^{t_{\rm max}} \frac{dE^{\pm}}{dt}
\]
\[
  = \int_{t_{\rm min}}^{t_{\rm max}} \int\,d\bx\, \left(p^{\pm} \frac{1}{\kappa}\frac{\partial p^{\pm}}{\partial t} + \bv^{\pm} \cdot \rho \frac{\partial \bv^{\pm}}{\partial t}\right)
\]
\[
  =\int_{t_{\rm min}}^{t_{\rm max}} \int\,d\bx\,\left(p^{\pm}(-\nabla \cdot \bv^{+} + h_s\delta(z-z_s)) + \bv^{\pm} \cdot \rho \frac{\partial \bv^{\pm}}{\partial t}\right)
\]
\[
=  \int_{t_{\rm min}}^{t_{\rm max}}\int\,d\bx\, \left(\left(\nabla p^{\pm}+\rho \frac{\partial \bv^{\pm}}{\partial t}\right) + p^{\pm}h_s\delta(z-z_s)\right)
\]
\begin{equation}
  \label{eqn:eident}
  =  \int_{t_{\rm min}}^{t_{\rm max}}\int\,dxdy h_s p^{\pm}|_{z=z_s} = \int_{t_{\min}}^{t_{\max}}\langle h_s(t), (\Lambda^{\pm}_{z_s} h_s)(t) \rangle_{L^2(\bR^2)}.
\end{equation}
That is, $\langle h_s(t), (\Lambda^{\pm}_{z_s} h_s)(t)
\rangle_{L^2(\bR^2)}$ is the rate of energy transfer from source to
the fluid at time $t$.

Assume that $h_s$ has compact support in time. Then $E^{\pm}
\rightarrow 0$ as $t \rightarrow \mp \infty$, and is eventually
constant as $t \rightarrow \pm \infty$. Then the energy identity
\ref{eqn:eident} implies that
\begin{equation}
  \label{eqn:etot}
  \pm \lim_{t \rightarrow \pm \infty} E^{\pm}(t) = \int \,dt\, \langle
  h_s(t), (\Lambda^{\pm}_{z_s} h_s)(t)\rangle_{L^2(\bR^2)} = \langle
  h_s,\Lambda^{\pm}_{z_s} h_s\rangle
\end{equation}
By virtue of identity \ref{eqn:ntodtr},
\[
  \lim_{t \rightarrow -\infty}E^{-}(t) =-\langle
  h_s,\Lambda^{-}_{z_s} h_s\rangle
\]
\[
  =  \langle  h_s,(\Lambda^{^+})^T_{z_s} h_s\rangle
\]
\[
  =  \langle  h_s,\Lambda^{^+}_{z_s} h_s\rangle
\]
so
\begin{equation}
  \label{eqn:surfenergy}
  = \lim_{t \rightarrow \infty}E^{+}(t)
\end{equation}
That is, the energy total energy transferred from the source to the
causal acoustic field as it evolves forward in time is the same as the
total energy transferred to the anti-causal acoustic field as it
evolves backwards in time.

Also onclude from identity \ref{eqn:etot} that the symmetric part of $\Lambda^+_{z_s}$ is positive semidefinite:
\begin{equation}
  \label{eqn:lamsd}
  \frac{1}{2}((\Lambda_{z_s}^+)^T + \Lambda^+_{z_s}) \ge 0.
\end{equation}

%Observe that the causal acoustic system \ref{eqn:awep} implies directly that
%\begin{equation}
%  \label{eqn:sourcevjump}
%  h_s= [v_z]|_{z=z_s}.
%\end{equation}

\section{Parametrix via Time Reversal}

Define $H$ to be the subspace $\{u \in (L^2(\bR^3))^4: \mbox{supp
}u\subset \{z<z_s\}\}$, by analogy with \cite{Lax:PDENotes},
p. 122. Note: must add microlocal $H^1$ constraint per
\cite[]{BaoSy:91b} so that trace operator is well-defined - postpone
this.

Given $u=(\bar{p}_0,\bar{\bv}_0) \in H$, extend in $t$ by solving the acoustic system
\begin{eqnarray}
\label{eqn:awe0}
  \frac{1}{\kappa}\frac{\partial p_0}{\partial t} & = & - \nabla \cdot \bv_0, \nonumber \\
  \rho\frac{\partial \bv_0}{\partial t} & = & - \nabla p_0,\nonumber \\
  p_0 & =& \bar{p}_0, t=0,\nonumber\\ 
  \bv_0 & = & \bar{\bv}_0, t=0,\nonumber\\
\end{eqnarray}
Note that $(p_0,\bv_0)=0$ near $t=0$ in the region $z>z_s$.

Denote by $P_s$, respectively $P_r$, the trace operator on $z=z_s$,
respectively $z=z_r$. Given $(p_0,\bv_0)$ as above, define
$(p_1,\bv_1)$ as the solution in $z \ne z_s$ of
\begin{eqnarray}
\label{eqn:awe1}
  \frac{1}{\kappa}\frac{\partial p_1}{\partial t} & = & - \nabla \cdot \bv_1, \nonumber \\
  \rho\frac{\partial \bv_1}{\partial t} & = & - \nabla p_1,\nonumber \\
  p_1 & =& 0,  \mbox{ for } 0,\nonumber\\ 
  \bv_1 & = & 0 \mbox{ for } 0,\nonumber\\
  P_sp_1 = P_sp_0. 
\end{eqnarray}
Extended to form a distribution on all of $\bR^4$, $(p_1,\bv_1)$ solve the
inhomogeneous system
\begin{eqnarray}
\label{eqn:awe1}
  \frac{1}{\kappa}\frac{\partial p_1}{\partial t} & = &
                                                        ((\Lambda^+_{z_s})^{-1}P_sp_0)\delta(z-z_s) - \nabla \cdot \bv_1, \nonumber \\
  \rho\frac{\partial \bv_1}{\partial t} & = & - \nabla p_1,\nonumber \\
  p_1 & =& 0,  \mbox{ for } t=0,\nonumber\\ 
  \bv_1 & = & 0 \mbox{ for } t=0.\nonumber\\
\end{eqnarray}
where
\begin{equation}
  \label{eqn:defhs}
  (\Lambda^+_{z_s})^{-1}P_sp_0 = P_s[(\bv_1)_z].
\end{equation}
